OpenMPIR

### 2017-09-08 paper.tex
Hacking away...

Consider the first code segment defined in \figref{search}.
Here, we find a simple OpenMP program which performs a divide-and-conquer search.
However, as written, there is a simple optimization that can be performed --
namely, that the value \code{(low+high)/2} may be computed once before either
parallel task, allowing the program to avoid some redundant computation. On
the serial version of this program, optimizations like this are performed
automatically. However, without the use of Tapir, these sorts of optimizations
are not run on parallel programs.

If one were to compile this program using the traditional technique of converting
to runtime calls, one would find a program similar to \figref{runtime_calls}.
The OpenMP pragmas are effectively treated as syntactic sugar for such runtime calls
because vanilla LLVM has no concise way to represent parallel programs.
These runtime calls obfuscate the program, making it infeasable to analyze or
optimize parallel code.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]{Simplified compilation of the unoptimized search code from \figref{search}.  \subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own functions and accompanying closures. These functions are then passed to OpenMP runtime calls. This system obfuscates the program, making it infeasable for an optimizer to reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without much modification.
To properly describe Tapir, one must briefly introduce key aspects of LLVM.
LLVM represents control flow through a control-flow graph (CFG) of blocks,
which are sequences of instructions. Blocks always end with terminator instructions
which define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
the conditional branch as it is also a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join frameworks.

@@

Initially, the cross-vendor OpenMP~\cite{spec} shared memory programming model 
focused on the execution of data parallelism by a cooperating team of threads, 
e.g., dividing the iterations of a loop among the threads. Version 3.0 of the 
OpenMP API specification introduced support for lightweight asynchronous tasks, 
designated by the application developer and scheduled onto the team of threads 
by the OpenMP run time implementation.  The \texttt{task} construct applied to 
a structured block of code creates an explicit task, and the \texttt{taskwait} 
construct waits for completion of all tasks generated by the current task.

@@

The initial design of the OpenMP task model~\cite{ayguade09design} established 
a basic framework for asynchronous task parallel execution in OpenMP programs.
Subsequent versions of the OpenMP specification up to the current version 4.5 
have added additional new features to the tasking model.  The \texttt{depend} 
clause codifies data dependences among tasks, indicating that a data location 
is an input or output of a task.  The run time system ensures that a task is 
not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
construct combines groups of independent loop iterations into explicit tasks, 
enabling composition of concurrent loop execution and independent explicit 
tasks within the same OpenMP parallel region. The \texttt{taskgroup} construct 
waits on not only all child tasks, but all descendent tasks, providing a deep 
synchronization. The \texttt{taskwait} construct allows the application to 
indicate a point at which the implementation may suspend the current task to 
work on other tasks, as may be desired for long-running tasks that generate 
many others.  The task concept was also leveraged to provide for asynchronous 
offload of data and computation to accelerators by applying the \texttt{nowait} 
clause to device constructs like the \texttt{target} construct to generate an 
asynchronous \textit{target task}.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:Example}
we show what Tapir code is generated.

@@

det.cont4:                                  
  sync label %sync.continue
\end{lstlisting}
 &  \\
\hline
\end{tabular}

\caption{Compilation to Tapir}
\label{fig:Example}
\end{figure*}

The listed Tapir LLVM IR code corresponds to lines 7 through 12 of the C source
code.  Upon entering the \texttt{else} branch, the code immediately detaches
the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in Figure~\ref{Fig:Example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.

>>>>

Consider the first code segment defined in~\figref{search}.
Here, we find a simple OpenMP program which performs a divide-and-conquer search.
However, as written, there is a simple optimization that can be performed --
namely, that the value \code{(low+high)/2} may be computed once before either
parallel task, avoiding redundant computations. For
the serial version of this program, optimizations like this are performed
automatically. However, without the use of Tapir, these sorts of optimizations
are not run on parallel programs.

If one were to compile this program using the traditional technique of lowering 
\texttt{\#pragma} directives into runtime calls, one would find a program similar 
to~\figref{runtime_calls}.
The OpenMP pragmas are effectively treated as syntactic sugar for runtime calls
because LLVM has no concise way to represent such parallel constructs.
Importantly, these runtime calls tend to obfuscate the program, making it 
infeasible to analyze and/or optimize parallel code.

@@

\caption[Simplified compilation of the unoptimized search code from \figref{search}.]
{Simplified compilation of the unoptimized search code from~\figref{search}.  
\subfigcap{a}~The function \code{search}. The parallel tasks are moved into their own 
functions and accompanying closures. These functions are then passed to OpenMP runtime 
calls. This system obfuscates the program, making it infeasable for an optimizer to 
reason about what is happening.}
  \label{fig:runtime_calls}
\end{figure}

\subsection{Tapir Internals}
Tapir is a compelling solution to this problem because it allows existing
analysis and optimizations to work on parallel programs without significant modifications.
To properly describe Tapir, one must briefly introduce key aspects of LLVM.
LLVM represents expression/evaluation order through a control-flow graph (CFG) of blocks that contain 
sequences of instructions. Blocks always end with terminator instructions that
define the edges of the control flow graph. For example, \figref{CFG} shows a simple
function and its corresponding LLVM code.

@@

To represent parallelism, Tapir introduces the \code{detach}, \code{reattach}, and
\code{sync} instructions. The \code{detach} statement is syntactically similar to
the conditional branch as it is also a terminator instruction with with two blocks.
However, unlike the conditional branch which denotes a choice between successors, the \code{detach} instruction denotes that its two successors may run in parallel. Semantically,
the first successor of a \code{detach} instruction is referred to as the \textbf{detached block} and represents a task that can, but is not required to, run in parallel with the
\textbf{continuation block}, or the second successor to the detach statement. The \code{detach}
is analagous to a \code{fork} in most fork-join models.

@@

Initially, the cross-vendor OpenMP~\cite{spec} shared memory programming model 
focused on the execution of data parallelism by a cooperating team of threads, 
e.g., dividing the iterations of a loop among the threads. Version 3.0 of the 
OpenMP API specification introduced support for lightweight asynchronous tasks, 
designated by the application developer and scheduled onto the team of threads 
by the OpenMP runtime implementation.  The \texttt{task} construct applied to 
a structured block of code creates an explicit task, and the \texttt{taskwait} 
construct waits for completion of all tasks generated by the current task.

@@

The initial design of the OpenMP task model~\cite{ayguade09design} established 
a basic framework for asynchronous task parallel execution in OpenMP programs.
Subsequent versions of the OpenMP specification up to the current version 4.5 
have added additional new features to the tasking model.  The \texttt{depend} 
clause codifies data dependences among tasks, indicating that a data location 
is an input or output of a task.  The runtime system ensures that a task is 
not scheduled until its input dependences are fulfilled.  The \texttt{taskloop} 
construct combines groups of independent loop iterations into explicit tasks, 
enabling composition of concurrent loop execution and independent explicit 
tasks within the same OpenMP parallel region. The \texttt{taskgroup} construct 
waits on not only all child tasks, but all descendent tasks, providing a deep 
synchronization. The \texttt{taskwait} construct allows the application to 
indicate a point at which the implementation may suspend the current task to 
work on other tasks, as may be desired for long-running tasks that generate 
many others.  The task concept was also leveraged to provide for asynchronous 
offload of data and computation to accelerators by applying the \texttt{nowait} 
clause to device constructs like the \texttt{target} construct to generate an 
asynchronous \textit{target task}.

@@

To better understand how the OpenMP to Tapir compiler works, we turn to the
\texttt{fib} example from Section~\ref{Sec:OpenMP}. In \figref{fig:tapir-example}
we show what Tapir code is generated.

@@

det.cont4:                                  
  sync label %sync.continue
\end{lstlisting}
 & \\
\hline
\end{tabular}
\caption{Transformation from OpenMP task constructs to Tapir.}
\label{fig:tapir-example}
\end{figure*}

The listed Tapir LLVM IR code corresponds to lines 7 through 12 of the C source
code.  Upon entering the \texttt{else} branch, the code immediately detaches
the basic block corresponding to the call to \texttt{fib(n-1)}, labeled
\texttt{det.achd}. The continuation for the first call also immediately
detaches the basic block corresponding to the call of \texttt{fib(n-2)}.
Finally, the continuation for the second call immediately calls sync. This is
actually a nice example of code that will be optimized by Tapir. The LLVM code
shown in Figure~\ref{Fig:tapir-example} is not optimized. With optimizataions turned
on, Tapir will replace a \texttt{detach} of a function call followed
immediately by \texttt{sync} with a simple function call, resulting in faster
code.
